---
title: "Section 5 MatchAlign"
output: html_document
date: '2022-06-24'
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

We load the main packages, including the R package
$\texttt{infinitefactor}$ that we developed.

```{r, message=FALSE, warning=FALSE}
rm(list = ls())
set.seed(1)
library(tidyverse)
library(infinitefactor)
library(coda)
library(factor.switching)
library(dplyr)
library(lavaan)
```

```{r}
source("helper_fcts.R")
```

We consider here the two applications from Section 4 of the paper.
The first concerns the measurements of $p=9$ mental ability test scores of $n=145$ seventh and eighth-grade students from the Grant-White School in Chicago (GrantWhite). 
In the second, we use data from the National Health and Nutrition
Examination Survey (NHANES) collected in 2015 and 2016. 
For NHANES, we select a subset of $107$ chemical exposures for which at least $30\%$ of
measurements have been recorded (i.e., are not missing). We also select
a subsample of $4468$ individuals for which at least $30\%$ of the $107$
chemicals have been recorded. After this preprocessing step, we are left
with a matrix $X$ of dimension $4468 \times 107$ with $36\%$ missing
data.

```{r}
which_data = 'NHANES' # %in% c('NHANES','GrantWhite')

if(which_data=='NHANES'){
  
  df <- readRDS("df_chem.rds")
  dim(df)

  k <- 25
  
} else if (which_data=='GrantWhite'){
  
  df <- HolzingerSwineford1939
  df <- df[df$school=="Grant-White",]
  df <- df %>% select(matches("^x\\d+$"))
  dim(df)

  k <- 3
  
} else {
  stop("Choose 'which_data' among c('NHANES','Grant-White')")
}
```

```{r}
df[1:10,1:7]
```

After pre-processing the data, some of the observations might still be missing (more than $35\%$ in NHANES).

```{r}
is.na(df) %>% mean()
```

We now inspect the eigen-decomposition of the correlation matrix, and we
noticed that the first $k$ eigenvectors explain more than $65\%$ of the
total variability in both cases.

```{r}
cor_X <- cor(df, use = "pairwise.complete.obs")
res <- eigen(cor_X) 
idx_pos <- res$values>0
eig_cumsum <- cumsum(res$values[idx_pos])/sum(res$values[idx_pos])
round( eig_cumsum[k] * 100, 2 )
```

## MSGP prior

We now estimate a Bayesian factor model using the MSGP prior, with with
$k$ latent variables.

We modified the function $\texttt{linear_MSGP}$ contained in the
$\texttt{infinitefactor}$ package in order to deal with missing data.
The new function $\texttt{linear_MSGP_NA}$ is contained in the file
$\texttt{helper_fcts.R}$.

Missing data inputation is straightforward in Bayesian factor model.
Given the values of $\Lambda$, $\{\eta_i\}_{i=1}^n$, and
$\Sigma=\text{diag}(\sigma_1^2,\dots,\sigma_p^2)$ at each iteration of
the MCMC chain, any missing entry $X_{ij}$ is sampled as
$$ X_{i j} \sim N(\lambda_j^\top \eta_i, \sigma_j^2) $$ where
$\lambda_j$ is the $j^{th}$ row of $\Lambda$.

```{r, cache=TRUE}
out <- linearMGSP_NA(log(df) %>% as.matrix() %>% scale(), verbose = F,
                     nrun = 10000, burn = 5000, thin = 5, 
                     kinit = k, adapt = F, output = "factSamples")
```

```{r}
lambdaSamps <- out$lambdaSamps
etaSamps <- out$etaSamps

lambda_fact_switch <- to_fact_switching(lambdaSamps)
```

## Post-Processing Algorithms

Now we run different post-processing algorithms, comparing MatchAlign
RSP and WOP. For each method we also compute also runtimes and ESS.

### WOP

```{r}
start_time <- Sys.time()
wop_ris <- quiet(weighted_procrustes_switching(lambda_mcmc = lambda_fact_switch, verbose = F,))
end_time <- Sys.time()
WOP_time <- end_time - start_time

Lambda_WOP <- to_infinitefact(wop_ris$lambda_reordered_mcmc, k, ncol(df))

WOP_ess <- apply(simplify2array(Lambda_WOP), c(1,2), effectiveSize) %>% mean()
```

### RSP-full-SA

```{r, cache=TRUE}
start_time <- Sys.time()
rspFull_ris <- quiet(rsp_full_sa(lambda_mcmc = lambda_fact_switch, sa_loops = 20, verbose = F))
end_time <- Sys.time()
rspFull_time = end_time - start_time

Lambda_rspFull <- to_infinitefact(rspFull_ris$lambda_reordered_mcmc, k, ncol(df))

rspFull_ess <- apply(simplify2array(Lambda_rspFull), c(1,2), effectiveSize) %>% mean()
```

### RSP-partial-SA

```{r, cache=TRUE}
start_time <- Sys.time()
rspPartial_ris <- quiet(rsp_partial_sa(lambda_mcmc = lambda_fact_switch, sa_loops = 20, verbose = F))
end_time <- Sys.time()
rspPartial_time = end_time - start_time

Lambda_rspPartial <- to_infinitefact(rspPartial_ris$lambda_reordered_mcmc, k, ncol(df))

rspPartial_ess <- apply(simplify2array(Lambda_rspPartial), c(1,2), effectiveSize) %>% mean()
```

### MatchAlign

```{r}
start_time <- Sys.time()
out_MatchAlign <- jointRot_NOnorm(lambdaSamps, etaSamps)
end_time <- Sys.time()

MatchAlign_time = end_time - start_time

Lambda_MatchAlign <- out_MatchAlign$lambda

MAl_ess <- apply(simplify2array(Lambda_MatchAlign), c(1,2), effectiveSize) %>% mean()
```

## Accuracy Metric

We further compute the accuracy metric (3.1):
$$ || \overline{\Lambda\Lambda^T \!\!}\,\, - \bar \Lambda_* \bar \Lambda_* ^T ||_F $$
where
$\overline{\Lambda\Lambda^T \!\!}\,\, = \frac{1}{T}\sum_{t = 1}^T \Lambda^{(t)}(\Lambda^{(t)})^T$
and $\bar \Lambda_* = \frac{1}{T} \sum_{t = 1}^T \Lambda^{(t)}_*$ is the
posterior mean of the aligned $\Lambda$.

In order to give a more direct intuition of the effective benefit of
each methodology, the metric values are expressed also as percentages of
the relative error in the posterior mean of the covariance matrix
$$ \frac{ || \overline{\Lambda\Lambda^T \!\!}\,\, - \bar \Lambda_* \bar \Lambda_*^T ||_F }{|| \overline{\Lambda\Lambda^T \!\!}\,\,\, ||_F} $$

```{r}
# samples
Lambda_LambdaT <- lapply(lambdaSamps, function(x) x %*% t(x))
Lambda_LambdaT_mean <- lmean(Lambda_LambdaT)

# Base line 
base_norm <- norm(Lambda_LambdaT_mean, type = "F")

# WOP
Align_cov <- lmean(Lambda_WOP)
Align_cov_mean <- Align_cov %*% t(Align_cov)
WOP_norm <- norm(Align_cov_mean - Lambda_LambdaT_mean, type = "F")

# RSP-full-SA
Align_cov <- lmean(Lambda_rspFull)
Align_cov_mean <- Align_cov %*% t(Align_cov)
rspFull_norm <- norm(Align_cov_mean - Lambda_LambdaT_mean, type = "F")

# RSP-partial-SA
Align_cov <- lmean(Lambda_rspPartial)
Align_cov_mean <- Align_cov %*% t(Align_cov)
rspPartial_norm <- norm(Align_cov_mean - Lambda_LambdaT_mean, type = "F")

# MatchAlign
LambdaAL_mean <- lmean(Lambda_MatchAlign)
Align_cov_mean <- LambdaAL_mean %*% t(LambdaAL_mean)
MatchAlign_norm <- norm(Align_cov_mean - Lambda_LambdaT_mean, type = "F")

# Unaligned
Align_cov <- lmean(lambdaSamps)
Align_cov_mean <- Align_cov %*% t(Align_cov)
Unaligned_norm <- norm(Align_cov_mean - Lambda_LambdaT_mean, type = "F")
```

## Runtimes

```{r}
print('Runtimes')
round(c(MatchAlign = MatchAlign_time, `RSP-full-SA` = rspFull_time,
  `RSP-partial-SA` = rspPartial_time,WOP=WOP_time),3)
```

## Accuracy Metric

```{r}
print('Absolute Accuracy Metric')
round(c(MatchAlign = MatchAlign_norm, `RSP-full-SA` = rspFull_norm,
  `RSP-partial-SA`=rspPartial_norm, WOP=WOP_norm, Unaligned=Unaligned_norm),3)
```

```{r}
print('Relative Accuracy Metric')
round(100*c(MatchAlign = MatchAlign_norm, `RSP-full-SA`=rspFull_norm,
            `RSP-partial-SA`=rspPartial_norm, WOP=WOP_norm, Unaligned=Unaligned_norm)/base_norm,1)
```

## ESS

```{r}
print('Absolute ESS')
round(c(MatchAlign=MAl_ess, `RSP-full-SA`=rspFull_ess,
        `RSP-partial-SA`=rspPartial_ess, WOP=WOP_ess),1)
```

```{r}
print('Relative ESS')
round(100*c(MatchAlign=MAl_ess, `RSP-full-SA`=rspFull_ess,
        `RSP-partial-SA`=rspPartial_ess, WOP=WOP_ess)/length(lambdaSamps),1)
```

## Plots

We now plot the matrix of factor loadings after using MatchAlign to
post-process the results. We re-arrange the rows of $\Lambda$ to make
the plot easier to interpret.

```{r}
mLambda_plot <- lmean(Lambda_MatchAlign)
ind_max = apply(mLambda_plot %>% abs(), 1, function(x) which(x == max(x))) %>%
  order()

plotmat(mLambda_plot[ind_max,])
```
