---
title: "Section 5 MatchAlign"
output: html_document
date: '2022-06-24'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

We load the main packages, including the R package *infinitefactor* that we developed. 

```{r, message=FALSE}
set.seed(1)
library(tidyverse)
library(infinitefactor)
library(coda)
library(factor.switching)
source("helper_fcts.R")
```

In this application, we use data from the National Health and Nutrition Examination Survey (NHANES) collected in 2015 and 2016. We select a subset of $107$ chemical exposures for which at least $30\%$ of measurements have been recorded (i.e., are not missing). We also select a subsample of $4468$ individuals for which at least $30\%$ of the $107$ chemicals have been recorded. After this preprocessing step, we are left with a matrix $X$ of dimension $4468 \times 107$ with $36\%$ missing data.


```{r}
# df_chem from Rds
df <- readRDS("data/df_chem_06_24.rds")
dim(df)
```

```{r}
head(df)
```

After pre-processing the data, more than $35\%$ of the observations are still missing. 

```{r}
is.na(df) %>% mean()
```

We now inspect the eigen-decomposition of the correlation matrix, and we noticed that the first $25$ eigenvectors explain more than $65\%$ of the total variability; hence, we set the number of factors equal to $25$.

```{r}
cor_X <- cor(df, use = "pairwise.complete.obs")
res <- eigen(cor_X) 
eig_cumsum <- cumsum(res$values)/sum(res$values)
round( eig_cumsum[25] * 100, 2 )
```

## MSGP prior

We now estimate a Bayesian factor model using the MSGP prior, with with $k = 25$ latent variables. To avoid running the full sampler, we recommend reading the file *out_06_24.rds* that contains the output of the function *linearMGSP_NA* with $k= 25$. We also provide the file *out_06_24_k_5.rds* which contains the result when the number of latent variables is equal to $5$.

We modified the function *linear_MSGP* contained in the *infinitefactor* package in order to deal with missing data. The details of our approach can be found in Section 5 of the paper, while the new function *linear_MSGP_NA* is contained in the file *helper_fcts.R*.


```{r}
k <- 25
out <- linearMGSP_NA(log(df) %>% as.matrix() %>% scale(),
                     nrun = 10000, burn = 5000, thin = 5,
              kinit = k, adapt = F, output = "factSamples")
```


```{r}
# out <- readRDS('data/out_06_24_k_5.rds'); k = 5
# out <- readRDS('data/out_25.rds'); k = 25
# out <- readRDS('data/out_40.rds'); k = 40
# str(out, 1)
```

This file contains 1000 posterior samples for $\Lambda$ and $\eta$.

```{r}
length(out$lambdaSamps)
dim(out$lambdaSamps[[1]])
```


## Post-Processing Algorithms

Now we run the MatchAlign post-processing algorithm:


```{r}
lambdaSamps <- out$lambdaSamps
etaSamps <- out$etaSamps

start_time <- Sys.time()
out_MatchAlign <- jointRot(lambdaSamps, etaSamps)
end_time <- Sys.time()

end_time - start_time
```

We can compare our method with rsp with partial simulated annealing. The two algorithms deal with the posterior samples of $\Lambda$ with two different data types, so we process the inputs and outputs to make them comparable.

```{r}
lambda_fact_switch <- to_fact_switching(lambdaSamps)

start_time <- Sys.time()
tmp <- quiet(rsp_partial_sa(lambda_mcmc = lambda_fact_switch, 
                            sa_loops = 5, 
                            verbose = F))
end_time <- Sys.time()
end_time - start_time
```

```{r}
Lambda_rsp <- to_infinitefact(tmp$lambda_reordered_mcmc, k, ncol(df))
Lambda_MatchAlign <- out_MatchAlign$lambda
```

We now compute metric (4.1): 
\begin{align*}
    || \overline{\Lambda\Lambda^T} - \bar \Lambda_* \bar \Lambda_* ^T ||_F
\end{align*}
where $\overline{\Lambda\Lambda^T} = \frac{1}{T}\sum_{t = 1}^T \Lambda^{(t)}(\Lambda^{(t)})^T$ and $\bar \Lambda_* = \frac{1}{T} \sum_{t = 1}^T \Lambda^{(t)}_*$ is the posterior mean of the aligned $\Lambda$. 


```{r}
# samples
Lambda_LambdaT <- lapply(lambdaSamps, function(x) x %*% t(x))
Lambda_LambdaT_mean <- lmean(Lambda_LambdaT)

# Match Align
LambdaAL_mean <- lmean(Lambda_MatchAlign)
Align_cov_mean <- LambdaAL_mean %*% t(LambdaAL_mean)
MatchAlign_norm <- norm(Align_cov_mean - Lambda_LambdaT_mean, type = "F")

# rsp full 
Align_cov <- lmean(Lambda_rsp)
Align_cov_mean <- Align_cov %*% t(Align_cov)
rsp_exact_norm <- norm(Align_cov_mean - Lambda_LambdaT_mean, type = "F")
```

```{r}
c(MatchAlign = MatchAlign_norm, rsp = rsp_exact_norm)
```

```{r}
plotmat(lmean(Lambda_rsp))
```

```{r}
plotmat(lmean(Lambda_MatchAlign))
```